---
title: "gRNA and iBAR calling for perturb-seq data"
author: "Magdalena Strauss"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes 
---

This script assignes gRNAs and iBAR barcodes to cells for the perturb-seq data.

# Setup 

```{r,output=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = c("pdf"),
    dpi=300
)
chooseCRANmirror(ind=69)
core_function_file <- "../../core_functions.R"
source(core_function_file)
library(dplyr)
library(stringr)
set.seed(12345)

folders_cellranger_gRNA <- c("../cellranger/CBE_gRNA/outs","../cellranger/ABE_gRNA/outs")
folders_cellranger_iBAR <- c("../cellranger/CBE_iBAR/outs","../cellranger/ABE_iBAR/outs")
file_name_gRNA_assignment <- "gRNA_assignment.rds"
file_name_iBAR_assignment <- "iBAR_assignment.rds" 
file_name_sce_list_after_QC <- "sce_list.rds"
file_name_sce_list_after_gRNA_iBAR_assignment <- "sce_list_gRNA_iBAR.rds"
ids <- c("CBE","ABE")
gRNA_number_file_CBE <- "validation_lib_CBE_with_gRNA_numbers.csv"
gRNA_number_file_ABE <- "validation_lib_ABE_with_gRNA_numbers.csv"
  
read_in_raw_matrix <- function(name_path)
{# for cellranger output
  matrix_dir = paste(name_path,"/raw_feature_bc_matrix/",sep="")
  barcode_path <- paste0(matrix_dir, "barcodes.tsv.gz")
  features_path <- paste0(matrix_dir, "features.tsv.gz")
  matrix_path <- paste0(matrix_dir, "matrix.mtx.gz")
  mat_raw <- readMM(file = matrix_path)
  feature_names = read.delim(features_path,
                             header = FALSE,
                             stringsAsFactors = FALSE)
  barcode_names = read.delim(barcode_path,
                             header = FALSE,
                             stringsAsFactors = FALSE)
  colnames(mat_raw) = barcode_names$V1
  rownames(mat_raw) = feature_names$V1
  return(mat_raw)
}

```


Read in the SingleCellExperiments after QC. 

```{r}
sce_list <- readRDS(file_name_sce_list_after_QC)
colnames(sce_list[[2]]) <- paste0(substring(colnames(sce_list[[2]]),1,17),"2")
```

Reading in the gRNA matrices.

```{r}
gRNA_matrices <- list()
for (k in 1:length(folders_cellranger_gRNA)){
  matrix_temp <- read_in_raw_matrix(folders_cellranger_gRNA[k])
  gRNA_matrices[[k]] <- matrix_temp[!(grepl("ENS",rownames(matrix_temp))),]
  colnames(gRNA_matrices[[k]]) <- paste0(substring(colnames(gRNA_matrices[[k]]),1,16),"-",toString(k))
  gRNA_matrices[[k]] <- gRNA_matrices[[k]][,colnames(gRNA_matrices[[k]])%in%colnames(sce_list[[k]])]
}
```

# gRNA assignment

We perform gRNA assignment using a mixture of skewed normal distributions as Cooper, Coelho, Strauss, et. al. (2023). https://doi.org/10.1101/2023.05.22.541777

```{r,fig.width=13,fig.height=7}
gRNA_assignment <- list()
for (j in 1:length(ids)){
  gRNA_assignment[[j]] <- barcode_calling( gRNA_matrices[[j]],paste0("gRNA ",ids[j]))
}
names(gRNA_assignment) <- ids
saveRDS(gRNA_assignment,file=file_name_gRNA_assignment)
```
Assigning gRNAs the SingleCellExperiment 

```{r}
sce_list_all_cells <- sce_list #sces with all cells, not only those assigned gRNAs

for (j in 1:2){
  sce_list[[j]]$gRNA <- NA
  sce_list[[j]]$iBAR <- NA
}


for (j in 1:length(sce_list)){
  xx <- intersect(colnames(sce_list[[j]]),names(gRNA_assignment[[j]]$barcode_assignment$barcode_assignment))
  sce_list[[j]] <- sce_list[[j]][,xx]
  sce_list[[j]]$gRNA <- gRNA_assignment[[j]]$barcode_assignment$barcode_assignment[xx]
}

for (j in 1:length(sce_list)){
 sce_list[[j]]$nr_gRNAs <- sapply(sce_list[[j]]$gRNA,function(x) str_count(x,"-")+1)
}


```

# iBAR assignment

Reading in the iBAR matrices

```{r}
iBAR_matrices <- list()
for (k in 1:length(folders_cellranger_iBAR)){
  matrix_temp <- read_in_raw_matrix(folders_cellranger_iBAR[k])
  iBAR_matrices[[k]] <- matrix_temp[!(grepl("ENS",rownames(matrix_temp))),]
  colnames(iBAR_matrices[[k]]) <- paste0(substring(colnames(iBAR_matrices[[k]]),1,16),"-",toString(k))
  iBAR_matrices[[k]] <- iBAR_matrices[[k]][,colnames(iBAR_matrices[[k]])%in%colnames(sce_list[[k]])]
}
```

As the noise levels for the gRNA umi counts are much lower then for the iBAR counts, we assign iBARs to cells with a unique gRNA assigned by using the iBAR with the highest umi count in the cell. Similarly, we assign n iBARs with the highest UMI counts to a cell with n gRNAs.

```{r}
assign_iBAR <- function(k,j){
  a <- names(sort(iBAR_matrices[[j]][,colnames(sce_list[[j]])[k]],decreasing=TRUE)[1:(sce_list[[j]]$nr_gRNAs[k])])
  if (length(a) > 1){
    b <- a
    a <- b[1]
    for (jk in 2:length(b)){
      a <- paste0(a,"-",b[jk])
    }
  }
  return(a)
}
for (j in 1:2){
  sce_list[[j]]$iBAR <- sapply(1:ncol(sce_list[[j]]),function(x) assign_iBAR(x,j))
}

```


Save the SingleCellExperiment. 

```{r}
saveRDS(sce_list,file_name_sce_list_after_gRNA_iBAR_assignment)
```

Numbers of cells called for each gRNA before and after QC. 

```{r}
gRNA_assignment <- readRDS(file_name_gRNA_assignment)
validation_lib <- read.table("../../validation_lib_annotated_230223.csv",sep=",",header=TRUE)
validation_lib_CBE <- validation_lib[validation_lib$editor=="CBE",]
validation_lib_ABE <- rbind(validation_lib[validation_lib$editor=="ABE",],
                           validation_lib[validation_lib$Gene=="N/A",])
validation_lib_CBE$nr_cells_after_QC <- gRNA_assignment$CBE$barcode_assignment$barcode_distribution_unique[validation_lib_CBE$sgRNA_ID]
validation_lib_ABE$nr_cells_after_QC <- gRNA_assignment$ABE$barcode_assignment$barcode_distribution_unique[validation_lib_ABE$sgRNA_ID]
validation_lib_ABE$nr_cells_after_QC[is.na(validation_lib_ABE$nr_cells_after_QC)] <- 0
validation_lib_CBE$nr_cells_after_QC[is.na(validation_lib_CBE$nr_cells_after_QC)] <- 0
write.table(validation_lib_CBE,file=gRNA_number_file_CBE,sep=",",col.names=TRUE,row.names=FALSE)
write.table(validation_lib_ABE,file=gRNA_number_file_ABE,sep=",",col.names=TRUE,row.names=FALSE)
```


