---
title: "Projection of diffusion score to drug addiction on CBE data set"
author: "Magdalena Strauss"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes 
---

# Projecting DA based diffusion score to CBE data set

```{r, echo=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = c("pdf"),
    dpi=300
)
library(scran)
library(scater)
library(ggplot2)
library(ggridges)
library(batchelor)
library(energy)
sce_ABE <- readRDS("sce_ABE.rds")
sce_CBE <- readRDS("sce_CBE.rds")
colourscheme_variant_class <- c("control"="lightgrey","canonical drug resistance"="darkblue","driver"="orange","drug addiction"="darkred", "NT_control" = "black")
```

We find for each resistance iBAR group in the CBE data set the 5 closest iBAR groups in the ABE data set, to transfer the diffusion scores for the transition to drug addiction (there are no drug addiction conferring gRNAs in the CBE data set, so we need to do this mapping), where closest refers to rank correlation between ABE and CBE based on the genes most rank-correlated with the diffusion score for the ABE data set. Then the CBE iBAR group is assigned the mean diffusion score of those five ABE iBAR groups.


```{r}
sce_ABE_resistance <- sce_ABE[,!(is.na(sce_ABE$diffusion_score))]
logcounts_temp <- logcounts(sce_ABE_resistance)
sds <- apply(logcounts_temp,1,sd)
logcounts_temp <- logcounts_temp[sds > quantile(sds,0.9),]
pval_cor_with_diffusion_score <- apply(logcounts_temp,1,function(x) cor.test(x,sce_ABE_resistance$diffusion_score,method="spearman")$p.value)
sig_genes <- names(pval_cor_with_diffusion_score)[p.adjust(pval_cor_with_diffusion_score,method="BH") < 10-6]
logcounts_resistance_sig <- logcounts_temp[sig_genes,]

logcounts_CDR_CBE <- logcounts(sce_CBE[sig_genes,sce_CBE$variant_class_highest_impact=="canonical drug resistance"])
cor_CDR_CBE <- cor(logcounts_CDR_CBE,logcounts_resistance_sig,method="spearman")
max_cor_CDR_CBE <- apply(cor_CDR_CBE,1,function(x) colnames(cor_CDR_CBE)[order(x,decreasing = TRUE)[1:5]])
diffusion_score_max_cor_CDR_CBE <- apply(max_cor_CDR_CBE,2,function(x) mean(sce_ABE_resistance[,x]$diffusion_score))
names(diffusion_score_max_cor_CDR_CBE) <- rownames(cor_CDR_CBE)
```

Now we plot the projected diffusion scores.

```{r}
validation_lib <- read.table("../../sc_HT29_debcet_annotated_270623.csv",sep=",",header=TRUE)
#replace the line above using supplementary table 1 from our preprint
feature_ref <- read.table("feature_ref_gRNA_CBE.csv",sep=",",header=TRUE)
validation_lib <- validation_lib[validation_lib$editor=="CBE" & validation_lib$sgRNA_ID%in%feature_ref$id,]

sce_CBE_resistance <- sce_CBE[,names(diffusion_score_max_cor_CDR_CBE)]
sce_CBE_resistance$diffusion_score_DA <-   diffusion_score_max_cor_CDR_CBE 

sce_CBE_resistance_one_gRNA <- sce_CBE_resistance[,!(grepl("-",sce_CBE_resistance$gRNA))]
gRNAs_resistance_CBE <- names(table(sce_CBE_resistance_one_gRNA$gRNA))[table(sce_CBE_resistance_one_gRNA$gRNA) >= 10]

df_gRNA_ds_mean <- unlist(lapply(1:length(gRNAs_resistance_CBE),function(x) mean(sce_CBE_resistance_one_gRNA$diffusion_score_DA[sce_CBE_resistance_one_gRNA$gRNA==gRNAs_resistance_CBE[x]])))
names(df_gRNA_ds_mean ) <- gRNAs_resistance_CBE

levels_a <- names(df_gRNA_ds_mean)[order(df_gRNA_ds_mean,decreasing=TRUE)]


df <- data.frame(diffusion_score_DA=sce_CBE_resistance_one_gRNA$diffusion_score_DA,gRNA=sce_CBE_resistance_one_gRNA$gRNA)

split_gRNAs <- function(line_df){
  output <- line_df
  if (sum(grepl("-",line_df$gRNA))>=1)
    {
    w <- strsplit(line_df$gRNA,"-")[[1]]
      for (k in 1:length(w)){
        output <- rbind(output,unlist(c(line_df[1],w[k])))
      }
    }
  return(output)
}

split_gRNAs_df <- function(d_f){
  output <- c()
  for (j in 1:nrow(d_f)){
    output <- rbind(output,split_gRNAs(d_f[j,]))  }
  output <- output[!(grepl("-",output$gRNA)),]
}

df <- split_gRNAs_df(df)
df <- df[df$gRNA%in%gRNAs_resistance_CBE,]


df$diffusion_score_DA <- as.double(df$diffusion_score_DA)
df$mean_diffusion_score_DA <- df_gRNA_ds_mean[as.vector(df$gRNA)]
df$variant_class <- validation_lib$variant_class[match(as.vector(df$gRNA),validation_lib$ID)]
df$gRNA=factor(df$gRNA,levels=levels_a)


sce_ABE_resistance_one_gRNA <- sce_ABE_resistance[,!(grepl("-",sce_ABE_resistance$gRNA))]
gRNAs_resistance_ABE <- names(table(sce_ABE_resistance_one_gRNA$gRNA))[table(sce_ABE_resistance_one_gRNA$gRNA) >= 10]


df_gRNA_ds_mean_ABE <- unlist(lapply(1:length(gRNAs_resistance_ABE),function(x) mean(sce_ABE_resistance_one_gRNA$diffusion_score[sce_ABE_resistance_one_gRNA$gRNA==gRNAs_resistance_ABE[x]])))

p2 <- ggplot(as.data.frame(df),aes(x=diffusion_score_DA,y=gRNA,fill=mean_diffusion_score_DA)) + geom_density_ridges() + scale_fill_viridis_c(option="magma",limits=c(min(df_gRNA_ds_mean_ABE),max(df_gRNA_ds_mean_ABE)))+
  theme_classic(base_size=24)+ylab("")+theme(legend.position = "none")
print(p2)
```
# Projected gRNA clusters for CBE data

Similarly to the diffusion scores, we also project the gRNA clusters to the CBE data set. Each CBE iBAR group is assigned the most frequent cluster among the cluster associations of the five most rank-correlated ABE iBAR groups. 

```{r}
library(stringr)
validation_lib_ABE <- read.table("validation_lib_ann_ABE.csv",sep=",",header=TRUE)
gRNA_max_cor_CDR_CBE <- apply(max_cor_CDR_CBE,2,function(x) sce_ABE_resistance[,x]$gRNA)
colnames(gRNA_max_cor_CDR_CBE) <- rownames(cor_CDR_CBE)
gRNA_max_cor_CDR_CBE <- gRNA_max_cor_CDR_CBE[,sapply(colnames(gRNA_max_cor_CDR_CBE),function(x) str_count(x,"-"))==1]
cluster_max_cor_CDR_CBE <- apply(gRNA_max_cor_CDR_CBE ,1,function(x) validation_lib_ABE$cluster[match(x,validation_lib_ABE$ID)])

find_mode <- function(v){
  table_v <- sort(table(v),decreasing=TRUE)
  if (length(table_v)==1){
    return(unique(v))
  }else{
      if (table_v[1] > table_v[2]){
      return(names(table_v)[1])
    }else{
      return(names(table_v)[sample(1:2,1)])
    }
  }
}

mapped_clusters <- apply(cluster_max_cor_CDR_CBE,1,find_mode)
names(mapped_clusters) <- colnames(gRNA_max_cor_CDR_CBE)

#averaging the mapped clusters across gRNAs
mapped_cluster_per_gRNA <- sapply(gRNAs_resistance_CBE,function(x) find_mode(mapped_clusters[grepl(x,names(mapped_clusters))]))
```

Finally, we add mapped diffusion scores and clusters to the annotated validation library. 

```{r}
validation_lib_ann_CBE <- read.table("validation_lib_ann_CBE.csv",header=TRUE,sep=",")
validation_lib_ann_CBE$mapped.cluster <- NA
validation_lib_ann_CBE$mapped.diffusion.score <- NA
validation_lib_ann_CBE$mapped.cluster[match(names(mapped_cluster_per_gRNA),validation_lib_ann_CBE$ID)] <- mapped_cluster_per_gRNA
validation_lib_ann_CBE$mapped.diffusion.score[match(names(df_gRNA_ds_mean),validation_lib_ann_CBE$ID)] <- df_gRNA_ds_mean
write.table(validation_lib_ann_CBE,file="validation_lib_ann_CBE.csv",row.names=FALSE,col.names=TRUE,sep=",")
sce_CBE$diffusion_score_DA <- NA
sce_CBE[,colnames(sce_CBE_resistance)]$diffusion_score_DA <- sce_CBE_resistance$diffusion_score_DA
saveRDS(sce_CBE,file="sce_CBE.rds")
```
