---
title: "QC for perturb-seq data"
author: "Magdalena Strauss"
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes 
bibliography: QC.bib
---

# QC by total umi count, number of features detected, and percentage of mitochondrial DNA

This notebook performs outlier-based QC for the non-genotyped large base editor screen, removing low outliers for the total count, low outliers for the number of detected features and high outliers for the percentage of counts from mitochondrial genes.

```{r,output=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = "pdf",
    dpi=300
)

library(Matrix)
library(ggplot2)
library(AnnotationHub)
library(AnnotationDbi)
library(scran)
library(scater)
library(grateful)
library(gridExtra)

folders_cellranger <- c("../cellranger/cellranger700_count_46539_7105STDY13259924_GRCh38-2020-A","../cellranger/cellranger700_count_46539_7105STDY13259926_GRCh38-2020-A")
file_name_mRNA_matrices <- "mRNA_matrices.rds"
file_name_mRNA_matrices_QC <- "mRNA_matrices_QC.rds"
file_name_sces <- "sce_list.rds"
ids <- c("CBE","ABE")

source("../../core_functions.R")
```


```{r}
mRNA_matrices <- list()
for (k in 1:length(folders_cellranger)){
 matrix_temp <- read_in_filtered_matrix(folders_cellranger[k])
  mRNA_matrices[[k]] <- matrix_temp[grepl("ENS",rownames(matrix_temp)),]
}
names(mRNA_matrices) <- ids
saveRDS(mRNA_matrices,file_name_mRNA_matrices)
```

Number of cells before QC: `r ncol(mRNA_matrices[[1]]) + ncol(mRNA_matrices[[2]])`
In the ABE data set: `r ncol(mRNA_matrices[[2]])`
In the CBE data set: `r ncol(mRNA_matrices[[1]])`

```{r}
discard <- list()
mRNA_matrices_QC <- list()
for (j in 1:length(mRNA_matrices)){
  discard[[j]] <- QC_mRNA_outlier(mRNA_matrices[[j]],file_name = paste0("Sample_",ids[j]))
  mRNA_matrices_QC[[j]] <- mRNA_matrices[[j]][,!(discard[[j]])]
}
names(mRNA_matrices_QC) <- ids
saveRDS(mRNA_matrices_QC,file=file_name_mRNA_matrices_QC)
```

Number of cells after QC: `r ncol(mRNA_matrices_QC[[1]]) + ncol(mRNA_matrices_QC[[2]])`
In the ABE data set: `r ncol(mRNA_matrices_QC[[2]])`
In the CBE data set: `r ncol(mRNA_matrices_QC[[1]])`

# Creating SingleCellExperiment and Normalisation

```{r}
set.seed(42)
getHVGs <- function(sce,min_mean = 0.001,FDR = 0.01)
{
  stats <- modelGeneVar(sce)
  stats <- stats[stats$mean > min_mean,]
  top_genes <- rownames(stats[stats$FDR < FDR,])
  return(top_genes)
}
sce_list <- list()
for (j in 1:2){
  sce_list[[j]] <- SingleCellExperiment(
    assays = list(counts = mRNA_matrices_QC[[j]]), colData = colnames(mRNA_matrices_QC[[j]]))
  clusts = as.numeric(quickCluster(sce_list[[j]], method = "igraph", min.size = 100))
  sce_list[[j]] = computeSumFactors(sce_list[[j]], clusters = clusts)
  sce_list[[j]] <- logNormCounts(sce_list[[j]])
}
names(sce_list) <- ids
saveRDS(sce_list,file=file_name_sces)
```



